---
title: "COMP7024 Programming for Data Science"
author: "B.M.D Amanda Rajapaksha (Student ID: 22216531)"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(kableExtra)
library(lubridate)
library(knitr)
library(ggplot2)

```

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or
damaged.
- We hereby certify that no part of this assignment/product has been copied from
any other student’s work or from any other source except where due acknowledgement
is made in the assignment.
- No part of this assignment/product has been written/produced for us by another
person except where such collaboration has been authorized by the subject lecturer/
tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection
software programs for the purpose of detecting possible plagiarism (which
may retain a copy on its database for future plagiarism checking).
- We hereby certify that we have read and understand what the School of Computing,
Engineering and Mathematics defines as minor and substantial breaches
of misconduct as outlined in the learning guide for this unit.

```{r}
orders <- read_csv("project_orders.csv")
products <- read_csv("project_product_info.csv")

head(orders)
head(products)
```


#template to answer the questions

## Question 1

### Explanation
The objective of carrying out this analysis is to assess the performance of the products by evaluating the total revenue generated by each product. This is achieved by using the orders data set, which contains information regarding each transaction in terms of the quantity of each product sold, the price at which it is being sold, etc. Additionally, the product data set contains descriptive information regarding each product. This information is linked together by matching the StockCode variable in each data set.

Before carrying out the analysis, it is important to clean the data to ensure that it is accurate. This is achieved by dropping rows that contain cancelled orders, as identified by the invoice column starting with the letter "C." Additionally, rows with negative quantities in the orders data set are removed. This is because negative values in the quantity column indicate invalid sales.

Revenue generated by each transaction is calculated by multiplying the quantity of each product by the price at which it is being sold. This is achieved by the formula Revenue = Quantity * Unit Price.

After that, the data is grouped by each product to calculate the total revenue generated by each product. This is achieved by ranking the products from highest to lowest revenue generated. Additionally, the percentage contribution of the top three products to the total revenue is calculated to assess the percentage contribution of each product to the total revenue generated by the organization.

### Code

```{r}
# Data Cleaning
orders_clean <- orders %>%
  filter(!str_detect(InvoiceNo, "^C")) %>%   # remove cancelled orders
  filter(Quantity > 0) %>%                   # remove negative quantities
  mutate(StockCode = toupper(StockCode))    # standardize stock codes

products <- products %>%
  mutate(StockCode = toupper(StockCode))

# Join datasets
data <- orders_clean %>%
  left_join(products, by = "StockCode")

# Remove unwanted descriptions
data <- data %>%
  filter(!Description %in% c("Adjust bad debt", "adjustment", "thrown away")) %>%
  filter(!is.na(Description))

# Calculate revenue
data <- data %>%
  mutate(Revenue = Quantity * UnitPrice)

# Revenue by product
product_revenue <- data %>%
  group_by(StockCode, Description) %>%
  summarise(TotalRevenue = sum(Revenue), .groups = "drop") %>%
  arrange(desc(TotalRevenue))

# Top 10 products
top10 <- product_revenue %>%
  slice_max(TotalRevenue, n = 10)

# Bottom 10 products
bottom10 <- product_revenue %>%
  slice_min(TotalRevenue, n = 10)

# Top 3 contribution
top3 <- product_revenue %>%
  slice_max(TotalRevenue, n = 3)

total_revenue <- sum(product_revenue$TotalRevenue)

top3_percentage <- sum(top3$TotalRevenue) / total_revenue * 100

# Display tables
kable(top10, caption = "Top 10 Products by Revenue")

kable(bottom10, caption = "Bottom 10 Products by Revenue")

top3_percentage
```

```{r}
#Visualisation
ggplot(top10, aes(x=reorder(Description,TotalRevenue), y=TotalRevenue))+
  geom_bar(stat="identity")+
  coord_flip()+
  labs(title="Top 10 Products by Revenue")
```

### Interpretation
The results show that there are products that yield significantly higher revenues than others. For instance, products like WHITE HANGING HEART T-LIGHT HOLDER and REGENCY CAKESTAND 3 TIER are among those that yield high revenues. This suggests that there is a small group of products that yield a significant amount of revenue. The revenue distribution is a normal occurrence in retail businesses, in which a small group of products is responsible for generating most of the revenue. From the graph, it is clear that a small percentage of products contribute significantly higher revenue compared to the rest. WHITE HANGING HEART T-LIGHT HOLDER generates the highest revenue, followed by DOTCOM POSTAGE and then REGENCY CAKESTAND 3 TIER.

This high difference between the top-generating products and the rest implies that revenue is generated by a small percentage of popular products that probably enjoy high and constant demand.

However, there is also a group of products that yield low or even no revenue at all. This could be due to low sales of these products, or it could be that these products are simply entries.

From a business point of view, these findings could be useful in:
- Identifying products that yield high revenue, which could be given priority in terms of promotional strategies
- Identifying products that yield low revenue, which could be given promotional strategies
- Identifying how revenue is being distributed among products in general.

## Question 2
###Explanation of the Task

The purpose of this study is to find the most valuable customers. The purchasing behavior of these customers is also analyzed. The customer value is calculated by determining the total revenue obtained from the transactions of these customers.

First, the transactions without a CustomerID are removed, as these cannot be attributed to any customer. Then, the total expenditure of each customer is calculated by adding up all the revenue obtained in their transactions.

Once the total expenditure of each customer is calculated, these customers are ranked in order to find the top five customers who spend the most. These customers form the most valuable customer group of the company.

Further analysis of the purchasing behavior of these customers is done by determining the products that these customers purchase the most. This is done by calculating the total quantity of these products purchased by these customers.

This method of analysis provides knowledge about the purchasing behavior of the most important customers of the company.


##Calculate Customer Spending
```{r}
# Step 1: Remove rows with missing CustomerID
data_clean <- data %>%
  filter(!is.na(CustomerID))

# Step 2: Calculate total spending per customer
customer_spending <- data_clean %>%
  group_by(CustomerID) %>%
  summarise(
    TotalSpending = sum(Revenue, na.rm = TRUE)
  ) %>%
  arrange(desc(TotalSpending))
```

```{r}
class(data)
ls()
```


```{r}
ls()

```


###Identify Top 5 Highest Spending Customer
```{r}
top_customers <- customer_spending %>%
  slice_head(n = 5)

# Display result
knitr::kable(top_customers, caption = "Top 5 Highest Spending Customers")

```
 
###Analyse Purchasing Behaviour of Top Customers
```{r}
top_customer_products <- data_clean %>%
  filter(CustomerID %in% top_customers$CustomerID) %>%
  group_by(CustomerID, Description) %>%
  summarise(
    TotalQuantity = sum(Quantity, na.rm = TRUE),
    .groups = "drop"
  )

```

###Identify top products purchased by each customer
```{r}
top_products_per_customer <- top_customer_products %>%
  group_by(CustomerID) %>%
  slice_max(TotalQuantity, n = 5) %>%
  arrange(CustomerID, desc(TotalQuantity))

# Display table
knitr::kable(top_products_per_customer,
             caption = "Top Purchased Products by High Value Customers")
```

```{r}
#Visualization

top_customer_products <- data_clean %>%
  filter(CustomerID %in% top_customers$CustomerID) %>%
  group_by(CustomerID, Description) %>%
  summarise(TotalQuantity = sum(Quantity), .groups="drop")

top_products_per_customer <- top_customer_products %>%
  group_by(CustomerID) %>%
  slice_max(TotalQuantity, n = 5)
library(ggplot2)

ggplot(top_products_per_customer,
       aes(x = reorder(Description, TotalQuantity),
           y = TotalQuantity,
           fill = factor(CustomerID))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top Purchased Products by High-Value Customers",
    x = "Product Description",
    y = "Total Quantity Purchased",
    fill = "Customer ID"
  ) +
  theme_minimal()
```
```{r}
# Check dataset structure
str(data_clean)

# Check number of unique customers
n_distinct(data_clean$CustomerID)

# Verify top customers
print(top_customers)
```

##Interpretation

The results show that a small number of customers generate a significant portion of the company's revenue. The top five customers have spent considerably more than the average customer, indicating that they are highly valuable to the business. By analysing their purchasing behaviour, it becomes clear that these customers frequently purchase specific products in large quantities. This suggests strong product preferences among high-value customers.
Further,  from the above bar chart, the products most commonly bought by the five highest-spending customers, based on the overall quantities bought, are shown. From the above chart, it is evident that the high-value customers are buying certain products in significantly higher quantities than the others. For example, PAPER CRAFT, LITTLE BIRDIE is bought in significantly higher quantities by the high-value customers compared to the quantities of the other products bought by them.

As can be understood from the above chart, the high-value customers may be bulk buyers or retailers who are looking to buy these products for resale purposes rather than for consumption purposes by the customers themselves, as they are buying the products in bulk quantities and not in smaller quantities for consumption purposes.

The fact that certain products are being bought by all the high-value customers is a clear indication of the high demand for these products by the high-value customers.

As can be understood from the above analysis, the high-value customers provide significant insight for the company, and the company can gain from the fact that by catering to the high-value customers, the overall stability of the revenue can be ensured.

#Q3
##Explanation
The objective of the present study is to observe the sales performance in terms of different countries and to identify the disparities in the region in terms of product consumption. Firstly, the total revenue generated in each country is calculated by adding the total revenue generated by each transaction associated with that country. This will help in identifying the markets that contribute to the overall sales. Further, the consumption of the product is analyzed by identifying the product that is most consumed in each country, based on the total quantity consumed. This will help in understanding the divergent behavior of consumers in each region. Incorporating revenue analysis with product demand analysis will help the organization attain a comprehensive understanding of the product

```{r}
# Calculate total revenue by country
country_revenue <- data %>%
  group_by(Country) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE)) %>%
  arrange(desc(TotalRevenue))

# Display table
kable(country_revenue, caption = "Total Revenue by Country")
# Calculate quantity sold for each product in each country
country_products <- data %>%
  group_by(Country, Description) %>%
  summarise(TotalQuantity = sum(Quantity, na.rm = TRUE), .groups = "drop")

# Identify most sold product in each country
top_product_country <- country_products %>%
  group_by(Country) %>%
  slice_max(TotalQuantity, n = 1)

# Display table
kable(top_product_country, caption = "Most Popular Product by Country")
```

```{r}
#Revenue Comparison Across Countries
ggplot(country_revenue,
       aes(x = reorder(Country, TotalRevenue),
           y = TotalRevenue)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Total Revenue by Country",
    x = "Country",
    y = "Total Revenue"
  ) +
  theme_minimal()
```
##Interpretation

The results show clear differences in revenue generation across countries. The United Kingdom generates the highest revenue, which is expected because the retailer is based in the UK and likely has a larger customer base there.

Other European countries such as Germany, France, and the Netherlands also contribute significantly to the company's revenue. This indicates that the retailer has a strong presence in the European market.

The analysis of product consumption shows that different countries have different product preferences. Some products are extremely popular in specific countries, suggesting that cultural preferences or seasonal demand may influence purchasing behaviour.

From a business perspective, these insights can help the company:

Identify high-performing international markets

Adjust marketing strategies for different countries

Promote products that are popular in specific regions

Expand operations in countries with strong demand

Understanding regional purchasing behaviour allows the business to better target customers and improve global sales performance.

#Q4
#(a) Explanation

The purpose of this task is to analyse sales trends over time using the transaction date variable InvoiceDate.

First, the month and year are extracted from the InvoiceDate variable to allow analysis of sales over time. This allows transactions to be grouped by month and year.

Next, the total revenue for each month is calculated by summing the revenue generated from all transactions in that month. This helps identify which months generate the highest sales.

In addition, the analysis identifies products that consistently appear among the lowest-selling products across multiple months. This helps determine which products perform poorly over time and may require review by the business.

Finally, the monthly revenue trend is visualised using a time-series line plot to observe any patterns or seasonal effects in sales.

```{r}
# Load libraries
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)

# Convert InvoiceDate to datetime and remove invalid dates
data <- data %>%
  mutate(InvoiceDate = dmy_hm(InvoiceDate)) %>%
  filter(!is.na(InvoiceDate))

# Create Year, Month, and YearMonth variables
data <- data %>%
  mutate(
    Year = year(InvoiceDate),
    Month = month(InvoiceDate),
    YearMonth = floor_date(InvoiceDate, "month")
  )

# Inspect
head(data)
colnames(data)

# Calculate monthly revenue
monthly_revenue <- data %>%
  group_by(YearMonth) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE),
            .groups = "drop")

# Plot monthly revenue trend
ggplot(monthly_revenue, aes(x = YearMonth, y = TotalRevenue)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Monthly Revenue Trend",
    x = "Month",
    y = "Total Revenue"
  ) +
  theme_minimal()

# Check range and summary
range(monthly_revenue$YearMonth)
summary(monthly_revenue$YearMonth)

```

```{r}
#Identify Highest Sales Months
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

kable(top_months, caption = "Months with Highest Sales")

# Identify top 5 highest revenue months
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

# Display
kable(top_months, caption = "Months with Highest Sales")
```

```{r}
# Calculate monthly revenue per product
monthly_product_sales <- data %>%
  group_by(Year, Month, Description) %>%
  summarise(
    MonthlyRevenue = sum(Revenue, na.rm = TRUE),
    .groups = "drop"
  )

# Identify lowest-selling product per month
lowest_products <- monthly_product_sales %>%
  group_by(Year, Month) %>%
  slice_min(MonthlyRevenue, n = 1, with_ties = FALSE)

# Display
kable(lowest_products, caption = "Lowest Selling Products per Month")

```

```{r}
# Optional: Monthly revenue trend with Year-Month interaction
monthly_revenue_interaction <- data %>%
  group_by(Year, Month) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE),
            .groups = "drop")

ggplot(monthly_revenue_interaction,
       aes(x = interaction(Year, Month),
           y = TotalRevenue,
           group = 1)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_point(color = "orange", size = 2) +
  labs(
    title = "Monthly Revenue Trend (Year-Month Interaction)",
    x = "Month",
    y = "Total Revenue"
  ) +
  theme_minimal()
```
#Interpretation

The monthly revenue analysis reveals fluctuations in sales across different months. Some months show significantly higher revenue compared to others, suggesting the presence of seasonal demand patterns.

For example, higher sales may occur during holiday seasons such as November and December, when customers tend to purchase more gifts and seasonal products. Lower sales months may occur during periods with less consumer spending.

The analysis of lowest-selling products across months indicates that some products consistently perform poorly. These products may have limited demand, outdated designs, or may not appeal to the majority of customers.

From a business perspective, these insights can help the company:

Identify seasonal sales trends

Improve inventory planning

Plan promotional campaigns during low-sales periods

Consider discontinuing or redesigning consistently low-performing products

Understanding sales trends over time allows the business to make better strategic decisions and optimise revenue throughout the year.

