---
title: "COMP7024 Programming for Data Science"
author: "B.M.D Amanda Rajapaksha (Student ID: 22216531)"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(kableExtra)
library(lubridate)
library(knitr)
library(ggplot2)
library(stringr)

```

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or
damaged.
- We hereby certify that no part of this assignment/product has been copied from
any other student’s work or from any other source except where due acknowledgement
is made in the assignment.
- No part of this assignment/product has been written/produced for us by another
person except where such collaboration has been authorized by the subject lecturer/
tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection
software programs for the purpose of detecting possible plagiarism (which
may retain a copy on its database for future plagiarism checking).
- We hereby certify that we have read and understand what the School of Computing,
Engineering and Mathematics defines as minor and substantial breaches
of misconduct as outlined in the learning guide for this unit.

```{r}
orders <- read_csv("project_orders.csv")
products <- read_csv("project_product_info.csv")

head(orders)
head(products)
```


#template to answer the questions

## Question 1

### Explanation
The objective of carrying out this analysis is to assess the performance of the products by evaluating the total revenue generated by each product. This is achieved by using the orders data set, which contains information regarding each transaction in terms of the quantity of each product sold, the price at which it is being sold, etc. Additionally, the product data set contains descriptive information regarding each product. This information is linked together by matching the StockCode variable in each data set.

Before carrying out the analysis, it is important to clean the data to ensure that it is accurate. This is achieved by dropping rows that contain cancelled orders, as identified by the invoice column starting with the letter "C." Additionally, rows with negative quantities in the orders data set are removed. This is because negative values in the quantity column indicate invalid sales.

Revenue generated by each transaction is calculated by multiplying the quantity of each product by the price at which it is being sold. This is achieved by the formula Revenue = Quantity * Unit Price.

After that, the data is grouped by each product to calculate the total revenue generated by each product. This is achieved by ranking the products from highest to lowest revenue generated. Additionally, the percentage contribution of the top three products to the total revenue is calculated to assess the percentage contribution of each product to the total revenue generated by the organization.

### Code

```{r}
# Data Cleaning
orders_clean <- orders %>%
  filter(!str_detect(InvoiceNo, "^C")) %>%   # remove cancelled orders
  filter(Quantity > 0) %>%                   # remove negative quantities
  mutate(StockCode = toupper(StockCode))    # standardize stock codes

products <- products %>%
  mutate(StockCode = toupper(StockCode))

# Join datasets
data <- orders_clean %>%
  left_join(products, by = "StockCode")

# Remove unwanted descriptions
data <- data %>%
  filter(!Description %in% c("Adjust bad debt", "adjustment", "thrown away")) %>%
  filter(!is.na(Description))

# Calculate revenue
data <- data %>%
  mutate(Revenue = Quantity * UnitPrice)

# Revenue by product
product_revenue <- data %>%
  group_by(StockCode, Description) %>%
  summarise(TotalRevenue = sum(Revenue), .groups = "drop") %>%
  arrange(desc(TotalRevenue))

# Top 10 products
top10 <- product_revenue %>%
  slice_max(TotalRevenue, n = 10)

# Bottom 10 products
bottom10 <- product_revenue %>%
  slice_min(TotalRevenue, n = 10)

# Top 3 contribution
top3 <- product_revenue %>%
  slice_max(TotalRevenue, n = 3)

total_revenue <- sum(product_revenue$TotalRevenue)

top3_percentage <- sum(top3$TotalRevenue) / total_revenue * 100

# Display tables
kable(top10, caption = "Top 10 Products by Revenue")

kable(bottom10, caption = "Bottom 10 Products by Revenue")

top3_percentage
```

```{r}
#Visualisation
ggplot(top10, aes(x=reorder(Description, TotalRevenue), y=TotalRevenue)) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  labs(
    title="Top 10 Products by Revenue",
    x="Product",
    y="Total Revenue"
  ) +
  theme_minimal()
```

### Interpretation
The results show that there are products that yield significantly higher revenues than others. For instance, products like WHITE HANGING HEART T-LIGHT HOLDER and REGENCY CAKESTAND 3 TIER are among those that yield high revenues. This suggests that there is a small group of products that yield a significant amount of revenue. The revenue distribution is a normal occurrence in retail businesses, in which a small group of products is responsible for generating most of the revenue. From the graph, it is clear that a small percentage of products contribute significantly higher revenue compared to the rest. WHITE HANGING HEART T-LIGHT HOLDER generates the highest revenue, followed by DOTCOM POSTAGE and then REGENCY CAKESTAND 3 TIER.

This high difference between the top-generating products and the rest implies that revenue is generated by a small percentage of popular products that probably enjoy high and constant demand.

However, there is also a group of products that yield low or even no revenue at all. This could be due to low sales of these products, or it could be that these products are simply entries.

From a business point of view, these findings could be useful in:
- Identifying products that yield high revenue, which could be given priority in terms of promotional strategies
- Identifying products that yield low revenue, which could be given promotional strategies
- Identifying how revenue is being distributed among products in general.

## Question 2
###Explanation of the Task

The purpose of this study is to find the most valuable customers. The purchasing behavior of these customers is also analyzed. The customer value is calculated by determining the total revenue obtained from the transactions of these customers.

First, the transactions without a CustomerID are removed, as these cannot be attributed to any customer. Then, the total expenditure of each customer is calculated by adding up all the revenue obtained in their transactions.

Once the total expenditure of each customer is calculated, these customers are ranked in order to find the top five customers who spend the most. These customers form the most valuable customer group of the company.

Further analysis of the purchasing behavior of these customers is done by determining the products that these customers purchase the most. This is done by calculating the total quantity of these products purchased by these customers.

This method of analysis provides knowledge about the purchasing behavior of the most important customers of the company.


##Calculate Customer Spending
```{r}
# Step 1: Remove rows with missing CustomerID
data_clean <- data %>%
  filter(!is.na(CustomerID))

# Step 2: Calculate total spending per customer
customer_spending <- data_clean %>%
  group_by(CustomerID) %>%
  summarise(
    TotalSpending = sum(Revenue, na.rm = TRUE)
  ) %>%
  arrange(desc(TotalSpending))
```

```{r}
class(data)
ls()
```


```{r}
ls()

```


###Identify Top 5 Highest Spending Customer
```{r}
top_customers <- customer_spending %>%
  slice_head(n = 5)

# Display result
knitr::kable(top_customers, caption = "Top 5 Highest Spending Customers")

```
 
###Analyse Purchasing Behaviour of Top Customers
```{r}
top_customer_products <- data_clean %>%
  filter(CustomerID %in% top_customers$CustomerID) %>%
  group_by(CustomerID, Description) %>%
  summarise(
    TotalQuantity = sum(Quantity, na.rm = TRUE),
    .groups = "drop"
  )

```

###Identify top products purchased by each customer
```{r}
top_products_per_customer <- top_customer_products %>%
  group_by(CustomerID) %>%
  slice_max(TotalQuantity, n = 5) %>%
  arrange(CustomerID, desc(TotalQuantity))

# Display table
knitr::kable(top_products_per_customer,
             caption = "Top Purchased Products by High Value Customers")
```

```{r}
#Visualization

top_customer_products <- data_clean %>%
  filter(CustomerID %in% top_customers$CustomerID) %>%
  group_by(CustomerID, Description) %>%
  summarise(TotalQuantity = sum(Quantity), .groups="drop")

top_products_per_customer <- top_customer_products %>%
  group_by(CustomerID) %>%
  slice_max(TotalQuantity, n = 5)
library(ggplot2)

ggplot(top_products_per_customer,
       aes(x = reorder(Description, TotalQuantity),
           y = TotalQuantity,
           fill = factor(CustomerID))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top Purchased Products by High-Value Customers",
    x = "Product Description",
    y = "Total Quantity Purchased",
    fill = "Customer ID"
  ) +
  theme_minimal()
```
```{r}
# Check dataset structure
str(data_clean)

# Check number of unique customers
n_distinct(data_clean$CustomerID)

# Verify top customers
print(top_customers)
```

##Interpretation

The results show that a small number of customers generate a significant portion of the company's revenue. The top five customers have spent considerably more than the average customer, indicating that they are highly valuable to the business. By analysing their purchasing behaviour, it becomes clear that these customers frequently purchase specific products in large quantities. This suggests strong product preferences among high-value customers.
Further,  from the above bar chart, the products most commonly bought by the five highest-spending customers, based on the overall quantities bought, are shown. From the above chart, it is evident that the high-value customers are buying certain products in significantly higher quantities than the others. For example, PAPER CRAFT, LITTLE BIRDIE is bought in significantly higher quantities by the high-value customers compared to the quantities of the other products bought by them.

As can be understood from the above chart, the high-value customers may be bulk buyers or retailers who are looking to buy these products for resale purposes rather than for consumption purposes by the customers themselves, as they are buying the products in bulk quantities and not in smaller quantities for consumption purposes.

The fact that certain products are being bought by all the high-value customers is a clear indication of the high demand for these products by the high-value customers.

As can be understood from the above analysis, the high-value customers provide significant insight for the company, and the company can gain from the fact that by catering to the high-value customers, the overall stability of the revenue can be ensured.

#Q3
##Explanation
The objective of this analysis is to evaluate the geographical distribution of the sales by determining the revenue generated and the demand for the products in different countries. This will provide valuable insights into the locations that generate the highest revenue and the popularity of the products. To do this, the transaction data set is used after eliminating the canceled orders and the invalid transactions. The revenue generated per transaction has already been computed as the product of Quantity and Unit Price. The data set is grouped based on the revenue generated per transaction, and the total revenue generated per country is computed. This will enable the determination of the countries that generate the highest revenue for the firm’s total sales. This will help in the evaluation of the geographical market concentration of the firm and determine whether the firm generates its revenue through domestic and international markets. Besides the revenue generated per transaction, the demand for the products is also evaluated by determining the most popular product sold in each country. This is achieved by grouping the data set based on the country and the description of the product, and the total quantity sold is computed.
The product with the highest quantity sold is determined as the most popular product sold in that country. The integration of these two evaluations will enable the determination of the macro-level and the micro-level analysis of the firm’s sales and the products sold in the different countries.

```{r}
# Calculate total revenue by country
country_revenue <- data %>%
  group_by(Country) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE)) %>%
  arrange(desc(TotalRevenue))

# Display table
kable(country_revenue, caption = "Total Revenue by Country")
# Calculate quantity sold for each product in each country
country_products <- data %>%
  group_by(Country, Description) %>%
  summarise(TotalQuantity = sum(Quantity, na.rm = TRUE), .groups = "drop")

# Identify most sold product in each country
top_product_country <- country_products %>%
  group_by(Country) %>%
  slice_max(TotalQuantity, n = 1)

# Display table
kable(top_product_country, caption = "Most Popular Product by Country")
```

```{r}
#Revenue Comparison Across Countries
ggplot(country_revenue,
       aes(x = reorder(Country, TotalRevenue),
           y = TotalRevenue)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Total Revenue by Country",
    x = "Country",
    y = "Total Revenue"
  ) +
  theme_minimal()
```
##Interpretation

The results indicate that the United Kingdom generates the highest total revenue by a wide margin, contributing the majority of total sales. This is expected since the retailer is based in the UK, meaning that it is expected that this market would contain the greatest number of customers with a higher purchasing activity. In addition, a number of other European countries, including EIRE, the Netherlands, Germany, and France, contribute a significant amount of revenue to total sales. This implies that the retailer has managed to expand beyond its domestic market and establish operations within the European region. The fact that these countries perform well implies that this market segment is a key contributor to international operations for this company. In contrast, a number of countries contribute a smaller amount of revenue compared to others. This can be attributed to a number of reasons, including a smaller number of customers or a lack of marketing within these regions.

Analysis of product demand in different countries indicates that different countries have different preferences for different products. Some products may be found to be top sellers in different countries, whereas some may be found to be highly popular in different regions. For instance, some products may be found to have particularly high demand in different regions of Europe. From a business strategy point of view, it is observed that such analyses are highly beneficial for business decision-making. By analyzing high-performing international markets, it is possible for the company to invest in these regions where high demand for their products exists. At the same time, analyzing regional product preferences may be highly beneficial for business in terms of promoting different products.

Overall, it is observed that the analysis highlights the significance of geographical sales patterns, which are highly beneficial for businesses in optimizing their marketing strategies in different regions of the world.

# Question 4
## Explaination

This study aims to examine the sales trends over time through the use of the InvoiceDate variable. The main purpose of this analysis is to identify the sales trends over the months and understand the performance of products in different months.To start the analysis, the month and year are extracted from the Invoice variable. This allows the sales transactions to be grouped according to the months in which they were made.  Consistent with the assumptions provided in the assignment, the sales transactions that were cancelled are not considered in the analysis. The cancelled sales transactions are identified as the ones where the invoice number starts with the letter "C." After the data cleansing process, the total revenue generated in each month is then computed. This allows the identification of the months where the highest sales are generated. Moreover, this analysis allows the identification of the months where the sales are fluctuating. Finally, a time series plot is then created to show the sales trends over the months. This allows the identification of the months where the sales revenue is augmented and diminished. Aside from the sales trends over the months, the performance of the products is also analyzed over the months. The analysis identified the products where the lowest sales revenue is generated over the months.
```{r}
# Load libraries
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)

# Convert date
data <- data %>%
  mutate(InvoiceDate = dmy_hm(InvoiceDate)) %>%
  filter(!is.na(InvoiceDate))

# Remove cancelled invoices
data <- data %>%
  filter(!grepl("^C", InvoiceNo))

# Calculate revenue
data <- data %>%
  mutate(Revenue = Quantity * UnitPrice)

# Create time variables
data <- data %>%
  mutate(
    Year = year(InvoiceDate),
    Month = month(InvoiceDate),
    YearMonth = floor_date(InvoiceDate, "month")
  )
#Monthly Revenue Calculation
monthly_revenue <- data %>%
  group_by(YearMonth) %>%
  summarise(
    TotalRevenue = sum(Revenue, na.rm = TRUE),
    .groups = "drop"
  )

#Highest Sales Months
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

kable(monthly_revenue, caption = "Monthly Revenue")
```



```{r}

#Identify Months with Highest Sales
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

kable(top_months, caption = "Top 5 Months with Highest Sales")

#Visualise Monthly Sales Trend (Time-Series Plot)
ggplot(monthly_revenue, aes(x = YearMonth, y = TotalRevenue)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Monthly Revenue Trend",
    x = "Month",
    y = "Total Revenue"
  ) +
  theme_minimal()

#Product Performance Over Time
monthly_product_sales <- data %>%
  group_by(YearMonth, Description) %>%
  summarise(
    MonthlyRevenue = sum(Revenue, na.rm = TRUE),
    .groups = "drop"
  )

#Identify Lowest-Selling Products Each Month
lowest_products <- monthly_product_sales %>%
  group_by(YearMonth) %>%
  slice_min(MonthlyRevenue, n = 1, with_ties = FALSE)

kable(lowest_products, caption = "Lowest-Selling Product Each Month")

#Products Consistently Among Lowest-Selling
consistent_low_products <- lowest_products %>%
  group_by(Description) %>%
  summarise(
    MonthsAsLowest = n()
  ) %>%
  arrange(desc(MonthsAsLowest))

kable(consistent_low_products, caption = "Products Consistently Among Lowest Sellers")
```

```{r}
#Identify Highest Sales Months
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

kable(top_months, caption = "Months with Highest Sales")

# Identify top 5 highest revenue months
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

# Display
kable(top_months, caption = "Months with Highest Sales")
```

```{r}
# Calculate monthly revenue per product
monthly_product_sales <- data %>%
  group_by(Year, Month, Description) %>%
  summarise(
    MonthlyRevenue = sum(Revenue, na.rm = TRUE),
    .groups = "drop"
  )

# Identify lowest-selling product per month
lowest_products <- monthly_product_sales %>%
  group_by(Year, Month) %>%
  slice_min(MonthlyRevenue, n = 1, with_ties = FALSE)

# Display
kable(lowest_products, caption = "Lowest Selling Products per Month")

```


```{r}
#Monthly revenue trend with Year-Month interaction
monthly_revenue_interaction <- data %>%
  group_by(Year, Month) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE),
            .groups = "drop")

ggplot(monthly_revenue_interaction,
       aes(x = interaction(Year, Month),
           y = TotalRevenue,
           group = 1)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_point(color = "orange", size = 2) +
  labs(
    title = "Monthly Revenue Trend (Year-Month Interaction)",
    x = "Month",
    y = "Total Revenue"
  ) +
  theme_minimal()
```
#Interpretation

The time-series analysis shows fluctuations in revenue for different months of the year. Some months show significantly higher revenue, possibly indicating a purchasing season for the products. In the past, retail businesses have shown increased sales in months such as November and December, as people buy gifts and seasonal products. Months with low consumer demand are those with low revenue in the analysis. The analysis has also shown products with consistently low sales in different months.

Some of the characteristics of these products could be:
-Low consumer demand for the products
-Lack of product visibility
-Outdated designs
-High price compared to similar products

As a manager, the analysis can help in:
-Improving inventory management
-Changing the marketing approach
-Launching promotions for products with low sales
-Removing products with consistently low sales

The revenue trajectory also implies that there may be some seasonal effects on sales, as some months show high demand for products. The data may be used by businesses to:

-increase stock levels before high-demand months
-create advertising campaigns during low-demand periods
-manage staffing and logistics during high-demand periods