---
title: "COMPXXXX Assignment"
author: "Your Name (Student ID)"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(kableExtra)
library(lubridate)
library(knitr)
library(ggplot2)

```

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or
damaged.
- We hereby certify that no part of this assignment/product has been copied from
any other student’s work or from any other source except where due acknowledgement
is made in the assignment.
- No part of this assignment/product has been written/produced for us by another
person except where such collaboration has been authorized by the subject lecturer/
tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection
software programs for the purpose of detecting possible plagiarism (which
may retain a copy on its database for future plagiarism checking).
- We hereby certify that we have read and understand what the School of Computing,
Engineering and Mathematics defines as minor and substantial breaches
of misconduct as outlined in the learning guide for this unit.

```{r}
orders <- read_csv("project_orders.csv")
products <- read_csv("project_product_info.csv")

head(orders)
head(products)
```


#template to answer the questions

## Q1
## (a)

### Explanation

The purpose of this analysis is to evaluate product performance based on the total revenue generated. The orders data contains information on transactions, such as the quantities and price per unit of each item, and the product information data contains product descriptions. These two datasets were joined based on the StockCode variable, enabling the addition of product descriptions to the orders data.

Before performing the analysis, the data was cleaned by deleting orders for canceled transactions and those with negative quantities, as they are deemed inaccurate for the purpose of the analysis. The revenue for each product was determined by multiplying the quantities of each item purchased by the price per unit of each item. The data was then split into groups based on the products, and the revenue generated by each product was determined.

The products were also ranked based on the revenue generated, and the best and worst-selling products were identified. The percentage contribution of the top three products to the revenue generated was also determined.  

### Code

```{r}
# Data Cleaning
orders_clean <- orders %>%
  filter(!str_detect(InvoiceNo, "^C")) %>%   # remove cancelled orders
  filter(Quantity > 0) %>%                   # remove negative quantities
  mutate(StockCode = toupper(StockCode))    # standardize stock codes

products <- products %>%
  mutate(StockCode = toupper(StockCode))

# Join datasets
data <- orders_clean %>%
  left_join(products, by = "StockCode")

# Remove missing descriptions
data <- data %>%
  filter(!is.na(Description))

# Calculate revenue
data <- data %>%
  mutate(Revenue = Quantity * UnitPrice)
# Revenue by product
product_revenue <- data %>%
  group_by(StockCode, Description) %>%
  summarise(TotalRevenue = sum(Revenue), .groups = "drop") %>%
  arrange(desc(TotalRevenue))

# Top 10 products
top10 <- product_revenue %>%
  slice_max(TotalRevenue, n = 10)

# Bottom 10 products
bottom10 <- product_revenue %>%
  slice_min(TotalRevenue, n = 10)

# Top 3 contribution
top3 <- product_revenue %>%
  slice_max(TotalRevenue, n = 3)
total_revenue <- sum(product_revenue$TotalRevenue)
top3_percentage <- sum(top3$TotalRevenue) / total_revenue * 100

# Display tables
kable(top10, caption = "Top 10 Products by Revenue")

kable(bottom10, caption = "Bottom 10 Products by Revenue")

top3_percentage
```

```{r}
#Visualisation
ggplot(top10, aes(x=reorder(Description,TotalRevenue), y=TotalRevenue))+
  geom_bar(stat="identity")+
  coord_flip()+
  labs(title="Top 10 Products by Revenue")
```

### Interpretation
This analysis represent the revenue generated from the sale of a number of products is significant. This indicates that revenue concentration exists, which means that the products that have generated the revenue are the ones that have performed better.

## Q2
#Calculate Customer Spending
```{r}
customer_spending <- data %>%
  group_by(CustomerID) %>%
  summarise(TotalSpending = sum(Revenue)) %>%
  arrange(desc(TotalSpending))

```
#Top 5 Customers
```{r}
top_customers <- head(customer_spending,5)
kable(top_customers)
```
 #Products Purchased by Top Customers
```{r}
top_customer_products <- data %>%
  filter(CustomerID %in% top_customers$CustomerID) %>%
  group_by(CustomerID, Description) %>%
  summarise(TotalQuantity = sum(Quantity)) %>%
  arrange(desc(TotalQuantity))
```
 # Visualization
```{r}
ggplot(top_customer_products,
       aes(x = reorder(Description, TotalQuantity),
           y = TotalQuantity,
           fill = factor(CustomerID))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top Purchased Products by Top Customers",
    x = "Product Description",
    y = "Total Quantity Purchased",
    fill = "Customer ID"
  ) +
  theme_minimal()
```
 

