---
title: "COMPXXXX Assignment"
author: "Your Name (Student ID)"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(kableExtra)
library(lubridate)
library(knitr)
library(ggplot2)

```

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or
damaged.
- We hereby certify that no part of this assignment/product has been copied from
any other student’s work or from any other source except where due acknowledgement
is made in the assignment.
- No part of this assignment/product has been written/produced for us by another
person except where such collaboration has been authorized by the subject lecturer/
tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection
software programs for the purpose of detecting possible plagiarism (which
may retain a copy on its database for future plagiarism checking).
- We hereby certify that we have read and understand what the School of Computing,
Engineering and Mathematics defines as minor and substantial breaches
of misconduct as outlined in the learning guide for this unit.

```{r}
orders <- read_csv("project_orders.csv")
products <- read_csv("project_product_info.csv")

head(orders)
head(products)
```


#template to answer the questions

## Q1
## (a)

### Explanation

The purpose of this analysis is to evaluate product performance based on the total revenue generated. The orders data contains information on transactions, such as the quantities and price per unit of each item, and the product information data contains product descriptions. These two datasets were joined based on the StockCode variable, enabling the addition of product descriptions to the orders data.

Before performing the analysis, the data was cleaned by deleting orders for canceled transactions and those with negative quantities, as they are deemed inaccurate for the purpose of the analysis. The revenue for each product was determined by multiplying the quantities of each item purchased by the price per unit of each item. The data was then split into groups based on the products, and the revenue generated by each product was determined.

The products were also ranked based on the revenue generated, and the best and worst-selling products were identified. The percentage contribution of the top three products to the revenue generated was also determined.  

### Code

```{r}
# Data Cleaning
orders_clean <- orders %>%
  filter(!str_detect(InvoiceNo, "^C")) %>%   # remove cancelled orders
  filter(Quantity > 0) %>%                   # remove negative quantities
  mutate(StockCode = toupper(StockCode))    # standardize stock codes

data <- data %>%
  filter(!Description %in% c("Adjust bad debt", "adjustment", "thrown away"))
products <- products %>%
  mutate(StockCode = toupper(StockCode))

# Join datasets
data <- orders_clean %>%
  left_join(products, by = "StockCode")

# Remove missing descriptions
data <- data %>%
  filter(!is.na(Description))


# Calculate revenue
data <- data %>%
  mutate(Revenue = Quantity * UnitPrice)
# Revenue by product
product_revenue <- data %>%
  group_by(StockCode, Description) %>%
  summarise(TotalRevenue = sum(Revenue), .groups = "drop") %>%
  arrange(desc(TotalRevenue))

# Top 10 products
top10 <- product_revenue %>%
  slice_max(TotalRevenue, n = 10)

# Bottom 10 products
bottom10 <- product_revenue %>%
  slice_min(TotalRevenue, n = 10)

# Top 3 contribution
top3 <- product_revenue %>%
  slice_max(TotalRevenue, n = 3)
total_revenue <- sum(product_revenue$TotalRevenue)
top3_percentage <- sum(top3$TotalRevenue) / total_revenue * 100

# Display tables
kable(top10, caption = "Top 10 Products by Revenue")

kable(bottom10, caption = "Bottom 10 Products by Revenue")

top3_percentage
```

```{r}
#Visualisation
ggplot(top10, aes(x=reorder(Description,TotalRevenue), y=TotalRevenue))+
  geom_bar(stat="identity")+
  coord_flip()+
  labs(title="Top 10 Products by Revenue")
```

### Interpretation
This analysis represent the revenue generated from the sale of a number of products is significant. This indicates that revenue concentration exists, which means that the products that have generated the revenue are the ones that have performed better.

## Q2
#Explanation of the Task

The goal of this task is to identify the most valuable customers for the business and analyse their purchasing behaviour.

First, the total spending of each customer is calculated by summing the revenue generated from all of their transactions. Revenue is defined as:
Revenue = Quantity * Unit price
Cancelled transactions (Invoice numbers starting with "C") must be excluded because they represent returned or cancelled purchases.

After calculating the total spending per customer, the Top 5 highest-spending customers are identified.

Then, for these customers, their purchasing behaviour is analysed by determining which products they purchased the most based on total quantity. This helps the business understand product preferences among its most valuable customers.


#Calculate Customer Spending
```{r}
# Step 1: Remove rows with missing CustomerID
data_clean <- data %>%
  filter(!is.na(CustomerID))

# Step 2: Calculate total spending per customer
customer_spending <- data_clean %>%
  group_by(CustomerID) %>%
  summarise(
    TotalSpending = sum(Revenue, na.rm = TRUE)
  ) %>%
  arrange(desc(TotalSpending))
```

```{r}
class(data)
ls()
```


```{r}
ls()

```


#Identify Top 5 Highest Spending Customer
```{r}
top_customers <- customer_spending %>%
  slice_head(n = 5)

# Display result
knitr::kable(top_customers, caption = "Top 5 Highest Spending Customers")

```
 
#Analyse Purchasing Behaviour of Top Customers
```{r}
top_customer_products <- data_clean %>%
  filter(CustomerID %in% top_customers$CustomerID) %>%
  group_by(CustomerID, Description) %>%
  summarise(
    TotalQuantity = sum(Quantity, na.rm = TRUE),
    .groups = "drop"
  )

```
#Identify top products purchased by each customer
```{r}
top_products_per_customer <- top_customer_products %>%
  group_by(CustomerID) %>%
  slice_max(TotalQuantity, n = 5) %>%
  arrange(CustomerID, desc(TotalQuantity))

# Display table
knitr::kable(top_products_per_customer,
             caption = "Top Purchased Products by High Value Customers")
```

```{r}
#Visualization

top_customer_products <- data_clean %>%
  filter(CustomerID %in% top_customers$CustomerID) %>%
  group_by(CustomerID, Description) %>%
  summarise(TotalQuantity = sum(Quantity), .groups="drop")

top_products_per_customer <- top_customer_products %>%
  group_by(CustomerID) %>%
  slice_max(TotalQuantity, n = 5)
library(ggplot2)

ggplot(top_products_per_customer,
       aes(x = reorder(Description, TotalQuantity),
           y = TotalQuantity,
           fill = factor(CustomerID))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top Purchased Products by High-Value Customers",
    x = "Product Description",
    y = "Total Quantity Purchased",
    fill = "Customer ID"
  ) +
  theme_minimal()
```
```{r}
# Check dataset structure
str(data_clean)

# Check number of unique customers
n_distinct(data_clean$CustomerID)

# Verify top customers
print(top_customers)
```

#Interpretation

The results show that a small number of customers generate a significant portion of the company's revenue. The top five customers have spent considerably more than the average customer, indicating that they are highly valuable to the business.

By analysing their purchasing behaviour, it becomes clear that these customers frequently purchase specific products in large quantities. This suggests strong product preferences among high-value customers.

From a business perspective, these customers are critical for maintaining revenue growth. The company could benefit from implementing loyalty programs, personalised promotions, or exclusive offers to retain these high-value customers. Additionally, ensuring that the products frequently purchased by these customers remain available in stock can help maintain customer satisfaction and continued spending.

#Q3
#(a) Explanation

The aim of this task is to analyse product consumption across different countries to understand global market demand.

First, the total revenue generated by each country is calculated. Revenue is defined as:
Revenue = Quantity * Unit price
The dataset already contains the calculated revenue variable from the previous task. Cancelled transactions were removed during the data cleaning stage, so only valid transactions are included in the analysis.

The data is grouped by Country to determine the total revenue generated in each country. This allows us to identify which countries contribute the most to overall sales.

Next, product consumption is analysed by identifying the single product with the highest total quantity sold in each country. This helps determine which products are most popular in different regions.

Finally, the results are presented in a summary table and visualised using a bar chart to compare revenue across countries.

```{r}
# Calculate total revenue by country
country_revenue <- data %>%
  group_by(Country) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE)) %>%
  arrange(desc(TotalRevenue))

# Display table
kable(country_revenue, caption = "Total Revenue by Country")
# Calculate quantity sold for each product in each country
country_products <- data %>%
  group_by(Country, Description) %>%
  summarise(TotalQuantity = sum(Quantity, na.rm = TRUE), .groups = "drop")

# Identify most sold product in each country
top_product_country <- country_products %>%
  group_by(Country) %>%
  slice_max(TotalQuantity, n = 1)

# Display table
kable(top_product_country, caption = "Most Popular Product by Country")
```

```{r}
#Revenue Comparison Across Countries
ggplot(country_revenue,
       aes(x = reorder(Country, TotalRevenue),
           y = TotalRevenue)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Total Revenue by Country",
    x = "Country",
    y = "Total Revenue"
  ) +
  theme_minimal()
```
#Interpretation

The results show clear differences in revenue generation across countries. The United Kingdom generates the highest revenue, which is expected because the retailer is based in the UK and likely has a larger customer base there.

Other European countries such as Germany, France, and the Netherlands also contribute significantly to the company's revenue. This indicates that the retailer has a strong presence in the European market.

The analysis of product consumption shows that different countries have different product preferences. Some products are extremely popular in specific countries, suggesting that cultural preferences or seasonal demand may influence purchasing behaviour.

From a business perspective, these insights can help the company:

Identify high-performing international markets

Adjust marketing strategies for different countries

Promote products that are popular in specific regions

Expand operations in countries with strong demand

Understanding regional purchasing behaviour allows the business to better target customers and improve global sales performance.

#Q4
#(a) Explanation

The purpose of this task is to analyse sales trends over time using the transaction date variable InvoiceDate.

First, the month and year are extracted from the InvoiceDate variable to allow analysis of sales over time. This allows transactions to be grouped by month and year.

Next, the total revenue for each month is calculated by summing the revenue generated from all transactions in that month. This helps identify which months generate the highest sales.

In addition, the analysis identifies products that consistently appear among the lowest-selling products across multiple months. This helps determine which products perform poorly over time and may require review by the business.

Finally, the monthly revenue trend is visualised using a time-series line plot to observe any patterns or seasonal effects in sales.

```{r}
# Load libraries
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)

# Convert InvoiceDate to datetime and remove invalid dates
data <- data %>%
  mutate(InvoiceDate = dmy_hm(InvoiceDate)) %>%
  filter(!is.na(InvoiceDate))

# Create Year, Month, and YearMonth variables
data <- data %>%
  mutate(
    Year = year(InvoiceDate),
    Month = month(InvoiceDate),
    YearMonth = floor_date(InvoiceDate, "month")
  )

# Inspect
head(data)
colnames(data)

# Calculate monthly revenue
monthly_revenue <- data %>%
  group_by(YearMonth) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE),
            .groups = "drop")

# Plot monthly revenue trend
ggplot(monthly_revenue, aes(x = YearMonth, y = TotalRevenue)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Monthly Revenue Trend",
    x = "Month",
    y = "Total Revenue"
  ) +
  theme_minimal()

# Check range and summary
range(monthly_revenue$YearMonth)
summary(monthly_revenue$YearMonth)

```

```{r}
#Identify Highest Sales Months
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

kable(top_months, caption = "Months with Highest Sales")

# Identify top 5 highest revenue months
top_months <- monthly_revenue %>%
  arrange(desc(TotalRevenue)) %>%
  slice_head(n = 5)

# Display
kable(top_months, caption = "Months with Highest Sales")
```

```{r}
# Calculate monthly revenue per product
monthly_product_sales <- data %>%
  group_by(Year, Month, Description) %>%
  summarise(
    MonthlyRevenue = sum(Revenue, na.rm = TRUE),
    .groups = "drop"
  )

# Identify lowest-selling product per month
lowest_products <- monthly_product_sales %>%
  group_by(Year, Month) %>%
  slice_min(MonthlyRevenue, n = 1, with_ties = FALSE)

# Display
kable(lowest_products, caption = "Lowest Selling Products per Month")

```

```{r}
# Optional: Monthly revenue trend with Year-Month interaction
monthly_revenue_interaction <- data %>%
  group_by(Year, Month) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE),
            .groups = "drop")

ggplot(monthly_revenue_interaction,
       aes(x = interaction(Year, Month),
           y = TotalRevenue,
           group = 1)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_point(color = "orange", size = 2) +
  labs(
    title = "Monthly Revenue Trend (Year-Month Interaction)",
    x = "Month",
    y = "Total Revenue"
  ) +
  theme_minimal()
```
#Interpretation

The monthly revenue analysis reveals fluctuations in sales across different months. Some months show significantly higher revenue compared to others, suggesting the presence of seasonal demand patterns.

For example, higher sales may occur during holiday seasons such as November and December, when customers tend to purchase more gifts and seasonal products. Lower sales months may occur during periods with less consumer spending.

The analysis of lowest-selling products across months indicates that some products consistently perform poorly. These products may have limited demand, outdated designs, or may not appeal to the majority of customers.

From a business perspective, these insights can help the company:

Identify seasonal sales trends

Improve inventory planning

Plan promotional campaigns during low-sales periods

Consider discontinuing or redesigning consistently low-performing products

Understanding sales trends over time allows the business to make better strategic decisions and optimise revenue throughout the year.

