---
author: Your Name (Student ID)
date: "`r Sys.Date()`"
output: html_document
title: COMPXXXX Assignment
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(kableExtra)
library(lubridate)
```
By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or
damaged.
- We hereby certify that no part of this assignment/product has been copied from
any other student’s work or from any other source except where due acknowledgement
is made in the assignment.
- No part of this assignment/product has been written/produced for us by another
person except where such collaboration has been authorised by the subject lecturer/
tutor concerned.
- We are aware that this work may be reproduced and submitted to plagiarism detection
software programs for the purpose of detecting possible plagiarism (which
may retain a copy on its database for future plagiarism checking).
- We hereby certify that we have read and understand what the School of Computing,
Engineering and Mathematics defines as minor and substantial breaches
of misconduct as outlined in the learning guide for this unit.
```{r}
orders <- read_csv("project_orders.csv")
products <- read_csv("project_product_info.csv")

head(orders)
head(products)
```


#template to answer the questions

## Q x
## (a)

### Explanation

The objective of this task is to prepare the data for analysis and examine product revenue performance.

Here, the Orders table is joined with the Product Info table using the StockCode variable so that each transaction includes a product description.

Next, cancelled transactions are removed because the assignment specifies that invoices starting with "C" represent cancellations and should not be included in revenue calculations.

Revenue is then calculated using:

Revenue = Quantity × UnitPrice

After calculating revenue, the total revenue for each product is aggregated. The Top 10 highest-selling products and Bottom 10 lowest-selling products are identified based on total revenue. Finally, the combined revenue contribution of the Top 3 products is calculated as a percentage of the overall revenue.

### Code

```{r}
# Remove cancelled transactions
orders_clean <- orders %>%
  filter(!str_starts(InvoiceNo, "C"))

# Join product information
data <- orders_clean %>%
  left_join(products, by = "StockCode")

# Calculate revenue
data <- data %>%
  mutate(Revenue = Quantity * UnitPrice)

# Revenue per product
product_revenue <- data %>%
  group_by(StockCode, Description) %>%
  summarise(TotalRevenue = sum(Revenue, na.rm = TRUE)) %>%
  arrange(desc(TotalRevenue))

# Top 10 products
top10 <- product_revenue %>% slice_max(TotalRevenue, n = 10)

# Bottom 10 products
bottom10 <- product_revenue %>% slice_min(TotalRevenue, n = 10)

# Top 3 contribution
top3 <- product_revenue %>% slice_max(TotalRevenue, n = 3)

total_revenue <- sum(product_revenue$TotalRevenue)

top3_percentage <- sum(top3$TotalRevenue) / total_revenue * 100

top10 %>% kable() %>% kable_styling()
bottom10 %>% kable() %>% kable_styling()

top3_percentage

```

```{r}
#Visualisation
ggplot(top10, aes(x=reorder(Description, TotalRevenue), y=TotalRevenue)) +
  geom_bar(stat="identity", fill="steelblue") +
  coord_flip() +
  labs(title="Top 10 Products by Revenue",
       x="Product",
       y="Revenue")
```

### Interpretation

Explain what the output means.

------------------------------------------------------------------------

I confirm that: - I understand all code submitted. - I can explain my
solution if required.
